This is the title [Article v\versionnumber]

### 2022-04-26 manuscript/manuscript.tex
Update manuscript.tex
solvation energy discussion

%%% Shortcuts and macros
\newcommand{\dgsolv}{\Delta G_\textup{solv}}
\newcommand{\software}{\emph}
\newcommand{\todo}{\textcolor{red}}
\newcommand\inlinecode{\texttt}

\lstset{
	basicstyle=\ttfamily\footnotesize
}
\lstdefinestyle{code}{}
\lstdefinestyle{python}{
	language=Python,
	stringstyle=\color[rgb]{1.0, 0, 0},
	commentstyle=\color[rgb]{0, 0, 0.6}
}
\lstset{style=code}

\DeclareSIUnit{\calorie}{cal}
\DeclareSIUnit{\kcalPerMolASqr}{\kilo\calorie\per\mole\per\angstrom\squared}

@@

\section{Introduction}
\todo{ (vah) Sources and better description of all these methods, more literature research, better flow }

Solvation thermodynamics govers any process involving solutes in a solvent.
Solutes are constantly interacting with their environment, necessitating a good description of these interactions for accurate computational predictions. 
Especially the interaction with water is of utmost importance, as many biological processes occur in a aquaeous environment. 
The free energy of solvation thereby influences real-world problems of interest such as hydrophobic effects, the binding of a ligand to a protein, protein-protein interactions, protein dynamics or even more generally, a protein's fold.
The computational calculation of thermodynamical solvation properties is therefore of particular interest and has been tackled by a wide array of different methods.   
Quantum mechanical methods provide the most rigorous description of the underlying potential energy surface (PES) but suffer from high computational cost, allowing only the treatment of small systems and few solvent molecules. 
More recently, machine learning approaches and QM/MM approaches are trying to bridge the gap to computationally less demanding methods while keeping the accuracy of ab initio QM methods.
Molecular mechanics based approaches tackle this problem by describing the underlying PES through empirically fit force-fields, which reduce the computational demand of these calculations tremendously.
This allows for the treatment of even large biological systems. 
Implicit solvent methods such as MM/PBSA  or MM/GBSA can quickly provide results for a large amount of supplied structures, but yield inaccurate results when compared to more rigorous explicit solvent methods. Somewhat inbetween the implicit solvent methods and the fully explicit solvent methods are Reference Interaction Site Models such as 3D-RISM. 
Starting with an explicit water model, 3D-RISM foregoes the explicit simulation and uses 3D solvent distributions based on the Ornstein--Zernike integral equations to predict solvation thermodynamics instead.
The 3D-RISM approach allows for a decomposition into entropic and enthalpic terms as well as a full localization of the underlying distributions to the solute structure.
The most statisticaly rigorous of the explicit solvent MM methods are alchemical methods such as FEP or TI. 
Assuming sufficient sampling, they can accurately predict the free energy of solvation in the boundaries of the assumptions underlying their forcefield. 
However, the seperation of enthalpic and entropic contributtions to the free energy of solvation is difficult with these methods and localizing these contributions not easily possible.
In this tutorial, the Grid Inhomogenous Solvation Theory (GIST) is used. It is based on Lazaridis' Inhomogenous Solvation Theory (IST), solving the arising integral equations on a grid. This allows for a full localization of entropic and enthalpic contributions on the grid around the solute. Recent improvements to the method have improved the calculation speed tremendously, allowing for the treatment of large macromolecules of biological interest such as proteases or antibodies. For small organic molecules, the method provides results compareable to TI in accuracy, provided some small correction factor is applied.

@@

\section{Prerequisites}
Before doing this tutorial, the reader should have a solid background knowledge of molecular dynamics (MD) simulations, as well as basic knowledge of Python, and should be familiar with working on the command line.

A recent version of \software{cpptraj} should be used, such as the one in AmberTools22 or an installation from GitHub.
Older versions of the Amber simulation engine can be used as well.
If no Amber installation is available, the MD can be performed using Gromacs.
In that case, we recommend preparing the structures and topologies using AmberTools and then converting to Gromacs using \software{acpype.py}.
In that way, a full Amber topology is available for the GIST analysis.

Furthermore, a recent Python version (>3.6) should be available with the following packages.

To run the MD simulation, a PC with recent GPU is recommended.
If the reader prefers to skip the MD and GIST calculation, we also provide the GIST output files with this tutorial, such that the post-processing can be done without any expensive calculations.

\begin{itemize}
	\item \software{gisttools} (and its prerequisites)
	\item \software{mdtraj}
\end{itemize}
%Here you would identify prerequisites/background knowledge that are assumed by your work, as well as any software/license requirements.

@@

\subsection{Solvation Energy}
\begin{equation}
	\Delta E_\textit{solv} = \Delta E_\textit{sw} + \Delta E_\textit{ww}
\end{equation}

\begin{equation}
	\Delta E_\textit{sw} = \rho^\textit{o} \int g_\textit{sw}\left(\textbf{r}\right) \Delta E_\textit{sw} \left(\textbf{r}\right) d\textbf{r}
\end{equation}

\begin{equation}
	\Delta E_\textit{ww} = \rho^\textit{o} \int g_\textit{sw}\left(\textbf{r}\right) \Delta E_\textit{ww} \left(\textbf{r}\right) d\textbf{r}
\end{equation}

@@

\subsection{Streptavidin/Biotin}
\subsection{Tutorial data}
We will use the 1STP crystal structure to start our calculations.
For the sake of simplicity, you can download prepared and solvated Amber topologies and structures based on 1STP by running
\begin{lstlisting}
git clone git@github.com:liedllab/gist-tutorial.git
\end{lstlisting}
This will download the tutorial files (including both the manuscript and the examples in the \software{code} folder).
\subsection{System Preparation and Equilibration}
If you want to skip this section, run 
\begin{lstlisting}
make equilibration-targets
\end{lstlisting}
in the code folder. This assumes that you have Python and Amber set up properly and that \software{pmemd.cuda} is in the PATH.

When calculating solvation free energy differences between a complex and the corresponding monomers (the dissociated state), there are two options for dealing with configurational diversity in the structures.
\todo{Maybe Valentin knows a citation for that?}
One can choose to sample the complex and dissociated states individually, or to sample only one of them and assume that there is sufficient overlap in the sampled conformations.
%One can choose to sample only the complex, assuming that all relevant conformations of the dissociated state are also found in the complex.
%The probability distribution of the dissociated state can be obtained by reweighting the probabilities in the complex using the solvation free energies.
%Alternatively, one can sample the complex and dissociated states individually.
%This approach does not require an overlap between the conformational probabilities of the complex and dissociated states.
%However, it is prone to statistical errors due to uncertainties in the ensembles, and requires the free energy contributions from the internal degrees of freedom of the monomers (in both the complex and the monomer states) to be known. \todo{This paragraph does not fit here well. Maybe put it somewhere else.}

Here, we choose not to include any sampling of the conformational states. 
It is therefore important that the conformations of the monomers in the complex and dissociated states match exactly. 
Use \software{cpptraj} to produce biotin and streptavidin structures based on the complex by stripping the respective other molecule like this:
\begin{lstlisting}
parm complex/solvated.parm7
trajin complex/equil/EQUIL-DONE.rst
strip ^2 parmout streptavidin/solvated.parm7
trajout streptavidin/pre-equil.ncrst
\end{lstlisting}
(Adapt for biotin as necessary.)

%We provide a script called \software{cpptraj\_remove\_mol.sh} that removes a single molecule from a topology and structure using \software{cpptraj} without altering the rest of the system. \todo{Explain cpptraj rather than our script. The script is for the automated workflow only.}

Use the \software{equilibration.py} script (or your own) to perform short NVT and NpT equilibration runs on the starting structure.
Note that the barostat might interfere with restrained equilibration of the complex, since different molecules are moved relative to each other when the volume changes.
But since our equilibration script also contains unrestrained pressure equilibration, we should not run into any trouble.

Then split the equilibrated structure into biotin and streptavidin parts using \software{cpptraj\_remove\_mol.sh}.
After that, you can use \software{equilibration.py -R} to run short restrained equilibrations (where only the water is allowed to move) on the individual systems.

\subsection{Running MD for GIST}

Now, you can run restrained MD simulations based on the equilibrated structures for subsequent GIST analyses.
We apply a harmonic restraint of \SI{100}{\kilo\calorie\per\mole\per\angstrom\squared} to all heavy atoms to keep the system in place.
We run \SI{100}{\nano\second} of simulation and store the conformations every \SI{100}{\pico\second}.
A simulation length of at least \SIrange{10}{20}{\nano\second} is recommended, to obtain statistically independent snapshots of the water.

An example Amber input file for GIST might look like this:
\begin{lstlisting}
restrained 100 ns NpT
&cntrl
  ntx=5, irest=1,
  ioutfm=1,
  ntb=2, iwrap=1,
  ntr=1, restraint_wt=100.0,
    restraintmask=\"!@H=&!:WAT\",
  ntp=1, pres0=1.0, taup=1.0,
  ntc=2, ntf=2,
  ntt=3, tempi=300.0, temp0=300.0, gamma_ln=2,
  nstlim=50000000, dt=0.002,
  ntwr=50000, ntwx=5000,
  ntpr=5000,
/
\end{lstlisting}
Note that these are very standard MD settings except for the line that defines the restraints.

Run the MD using \software{pmemd.cuda}:
\begin{lstlisting}
pmemd.cuda \
	-O \
	-i 100ns-npt-restraint.in \
	-o md-01.out \
	-p solvated.parm7 \
	-c EQUIL-DONE.rst \
	-x md-01.nc \
	-r md-01.ncrst \
	-ref EQUIL-DONE.rst
\end{lstlisting}
\subsection{Running the GIST analysis}

For easier post-processing, we will use grids that contain the whole molecule.
This could in theory be avoided for streptavidin and the complex, since only the surrounding of the biotin molecule (i.e., the binding pocket) is required for the analysis.
We also recommend that you center the molecule in space.
For each system (biotin, streptavidin, complex), produce a centered pdb file using \software{cpptraj}:
\begin{lstlisting}
parm solvated.parm7
trajin md-01.nc 1 1 1
autoimage !(:WAT) origin
strip :WAT
trajout solute-centered.pdb
\end{lstlisting}
Then, find the extent (minimum and maximum) of the x-, y-, and z-coordinates for each system.
This can be done e.g., using the \software{mdtraj} package in Python.
\begin{lstlisting}[style=python]
import mdtraj as md
pdb = md.load("solute-centered.pdb")
# mdtraj converts to nm.
# *10 to convert back to Angstrom.
min = pdb.xyz[0].min(0) * 10.
max = pdb.xyz[0].max(0) * 10.
print(*list(zip("xyz", min, max)), sep="\n")
\end{lstlisting}
Then, decide on grid dimensions for every system.
For a system centered at the origin with a spacing of \SI{0.5}{\angstrom} and a buffer of, e.g., \SI{7}{\angstrom} between the solute and the grid wall, you need $4.0\times (d+7)$ bins in the respective direction, where $d$ is the larger of $-$min and $+$max in the respective direction.
We note that this centered grid is usually larger than it needs to be in some directions.
If you want to save some extra memory and calculation time, feel free to align the grid more precisely to the minimum and maximum coordinates, or use a smaller grid that only covers the binding pocket.
You can also rotate the molecule by its principal axes to fit the cuboid grid more exactly.
However, this needs to be done \emph{before} the MD run, since rotating the trajectory damages the periodic box information.

Now, you can finally run the actual GIST analysis. Center the molecule exactly in the same way as above, to make sure that its position matches the grid parameters that you decided on.
We recommend using either the PME or GPU implementation of GIST, depending on whether you have a fast GPU.
A further advantage of the PME implementation is that it matches the energy calculation of the MD engine more closely.
\todo{At the time of writing, it is important to use an up-to-date version of PME-GIST, or the Esw-Eww decomposition will not be available. Maybe remove this for the final version, when Amber22 is out.}
\begin{lstlisting}
parm solvated.parm7
trajin md-01.nc
autoimage !(:WAT) origin
# remove the "pme" option if you want to use the GPU
gist griddim 100 100 100 out gist.dat \
  refdens 0.03287 pme
# other settings you might want to change
# (those are the defaults)
# gridcntr 0 0 0 gridspacn 0.5 temp 300
\end{lstlisting}
\subsection{Reference Values and Radial Convergence}
All GIST quantities should be expressed relative to bulk. In practice, this is not problematic for the entropies, since they are computed relative to a distribution of randomly oriented molecules at the reference density.
The solute-solvent energy Esw is also no problem, because it naturally tends to zero in bulk. 
However, the solvent-solvent energy Eww needs to be referenced.

The Amber manual provides reference energies for several solvents.
However, the exact reference value is different when using PME-GIST, and also depends on the box size.
For exact analyses such as computing the total $\dgsolv$ of a compound, it is important to compute an exact reference energy.

The most accurate method to compute a reference energy is to run a separate GIST calculations with the same energy method (PME/GPU) and a similar box size, and then compute the average solvent-solvent energy per molecule.

\begin{equation}
\label{eq_ewwref}
E_{ww}^\textit{ref} = \frac{\int{E_{ww}} \mathrm{d}x}{\left<N_\textit{solvent}\right>}
  = \frac{\sum_\textit{voxels}E_{ww}^\textit{dens} V_\textit{vox}}{\sum_\textit{voxels}{N_\textit{solvent}} / N_\textit{frames}}
\end{equation}

Alternatively, one can note that the solvent-solvent energy in a sufficiently large GIST grid converges to the correct bulk value at large distance to the solute.
This means that the reference energy can be obtained from a large GIST grid by binning the voxels by their distance to the solute, and evaluate Equation \ref{eq_ewwref} within each bin.
If this value converges to a constant, this can be used as $E_{ww}^\textit{ref}$.

\software{gisttools} contains functionality to perform this analysis easily.
There is also a method \inlinecode{detect\_reference\_value} that automatically tries to find the converged value fo the rdf.
Although it has a simple convergence check built-in, it is always recommended to check the convergence by hand as well.

\begin{lstlisting}[style=python]
import gisttools as gt
import matplotlib.pyplot as plt
gist = gt.gist.load_gist_file(
    "gist.dat", struct="solute-centered.pdb")
bins, eww = gist.rdf(
    "Eww_unref", bins=100, rmax=20., normalize="norm")
plt.plot(bins, eww)
eww_auto = gist.detect_reference_value().values[0]
plt.axhline(eww_auto, color='k')
plt.gca().set(
    xlabel='distance [A]', ylabel='E_ww per mol')
plt.show()
\end{lstlisting}

The expected output of this code is shown in Figure \ref{fig_ewwref}.

\begin{figure}
	\centering
	\includegraphics[width=0.8\linewidth]{figures/Eww_convergence.eps}
	\caption{Convergence of $E_{ww}$ with increasing distance to the solute. The horizontal line shows the automatically computed reference energy. Computed from the complex calculation.}
	\label{fig_ewwref}
\end{figure}

After choosing an appropriate reference value, we need to subtract this value from Eww.
Make sure to subtract the reference from the normalized (\_norm) data.
In \software{gisttools}, this is done simply by setting \inlinecode{gist.eww\_ref}.
In \software{gistpp}, this can be done as follows: \todo{This might become invalid with future GIST versions, depending on whether we adopt the COM as molecular center.}

\begin{lstlisting}
rho0=0.03287465
neg_eww_ref=9.533
gistpp -op multconst -i gist-gO.dx -opt const $rho0 \
    -o gist-gO-abs.dx 
gistpp -op div -i gist-Eww-dens.dx -i2 gist-gO-abs.dx \
    -o gist-Eww-norm.dx
gistpp -op addconst -i gist-Eww-norm.dx \
    -opt const $neg_eww_ref -o gist-Eww-ref-norm.dx
gistpp -op mult -i gist-Eww-ref-norm.dx \
    -i2 gist-gO-abs.dx -o gist-Eww-ref-dens.dx
\end{lstlisting}

Now, add all free energy contributions to obtain the spatially resolved $\dgsolv$.
In \software{gisttools}, this is done automatically (simply use the \inlinecode{A\_dens} and \inlinecode{A\_norm} columns).
In \software{gistpp}, use the \inlinecode{add} command, with the same syntax as the \inlinecode{div} command used above.
After that, check whether your free energy contributions become negligible at high distances to the solute.
In a plot like Figure \ref{fig_ewwref} based on the \inlinecode{A\_dens} column, the values should tend to zero.
However, it is more informative to plot the cumulative free energy contribution against the distance to the solute, to check the convergence.
It the curve flattens out, this is your final $\dgsolv$.
If it does not, you might need to tweak the Eww reference, or introduce a reference value for the entropy (especially with older versions of \software{cpptraj}).

\begin{lstlisting}[style=python]
import gisttools as gt
import matplotlib.pyplot as plt
# adapt eww_ref!
gist = gt.gist.load_gist_file("gist.dat",
    struct="solute-centered.pdb", eww_ref=-9.533)
bins, (dg, eww, esw, s) = gist.rdf(
    ["A_dens", "Esw_dens", "Eww_dens", "dTSsix_dens"],
    bins=100, rmax=20., normalize="none")
plt.plot(bins, np.cumsum(dg), label="dG")
plt.plot(bins, np.cumsum(eww), label="Eww")
plt.plot(bins, np.cumsum(esw), label="Esw")
plt.plot(bins, np.cumsum(s), label="dS")
\end{lstlisting}

The expected output of this code is shown in Figure \ref{fig_radial_convergence}.
In this case, both the Eww energy and the entropy are not perfectly flat, indicating that the reference value is not optimal.
The slow energy convergence indicates that a non-optimal reference value was used, but is also complicated by the fact that a charged solute is used without neutralizing counter ions.

\begin{figure}
	\centering
	\includegraphics[width=0.8\linewidth]{figures/A_E_S_convergence.eps}
	\caption{Convergence of $\dgsolv$ and its contributions with increasing distance to the solute. All quantities cumulative, i.e., summed up to the respective radius. Computed from the biotin calculation.}
	\label{fig_radial_convergence}
\end{figure}

\subsection{Visualizing $\dgsolv$}
Next, we want to visualize the 3-dimensional contributions to the free energy of hydration, as well as the entropy and enthalpy contributions.
You can use PyMol\cite{pymol} or VMD\cite{vmd} to visualize the .dx files from GIST.
With \software{gistpp}, you can use the file from the convergence check. Additional OpenDX files can be produced using \inlinecode{gistpp -op makedx}.
With \software{gisttools}, you can create one using e.g., \inlinecode{gist.save\_dx("A\_dens", "A\_dens.dx")}.

Visualize the free energy, the total energy, and the entropy, at several iso-levels.
The expected result is shown in Figure ???.

\subsection{Contribution of Hydration to Binding}

>>>>

%%% Shortcuts and macros
\newcommand{\dgsolv}{\Delta G_\textup{solv}}
\newcommand{\software}{\emph}
\newcommand{\todo}{\textcolor{red}}

\newenvironment{code}{\par \noindent \ttfamily}{}

@@

\section{Introduction}

@@

\section{Prerequisites}

%Here you would identify prerequisites/background knowledge that are assumed by your work, as well as any software/license requirements.

@@

\subsection{Solvation Energy}
The solvation energy is readily calculated by summing up the contributions of the water-water interaction and the water-solute interaction.
\begin{equation}
	\Delta E_\textit{solv} = \Delta E_\textit{sw} + \Delta E_\textit{ww}
\end{equation}

\begin{equation}
	\Delta E_\textit{sw} \equiv \frac{1}{8\pi^2} \int g_\textit{sw}\left(\omega|\textbf{r}\right) U_\textit{sw}\left(\textbf{r}, \omega\right) d\omega
\end{equation}

Similar to the entropy integrals, the solute-water solvation energy integral also decays with increasing distance from the solute. Hence, the solvation energy can be approximated by local spatial integrals. However, there is a complication arising from this approximation. Water molecules outside the region of interest will also be accounted for in the pairwise water term. For example, given two different regions $R_a$ and $R_b$, will not sum up to the total energy of the two regions due to double counting.

\begin{equation}
\begin{aligned}
	\Delta E_\textit{ww} \equiv & \left(\frac{1}{8\pi^2} \right)^2 \rho^o \int g_\textit{sw}\left(\omega|\textbf{r}\right) \\
	& \times \left[g_\textit{sw}\left(\textbf{r}^\prime , \omega^\prime \right) - g^o_\textit{ww} \left(\textbf{r}, \omega, \textbf{r}^\prime, \omega^\prime \right)\right] \\
	& \times U_\textit{ww}\left(\textbf{r}, \omega, \textbf{r}^\prime, \omega^\prime\right)d\omega d\textbf{r}^\prime d\omega^\prime
\end{aligned}
\end{equation}
There are two ways in which this double counting was accounted for in the original paper. First, using the water-displacement method. This method is applicable when a determining the energetic cost of kicking out a water molecule. 
\begin{equation}
	\Delta E^R_\textit{ww} = n^\textit{R} E^\textit{bulk}_\textit{ww} - E^\textit{R,corr}_\textit{ww}
\end{equation}

The second method uses normalized-water properties to account the difference in water properties of two different regions. The normalized water property is defined as the water property per water molecule. 
\begin{equation}
	\Delta E^\textit{R,norm}_\textit{ww} = E^\textit{R,norm}_\textit{ww} - 2E^\textit{bulk}_\textit{ww}
\end{equation}

@@

\subsection{Streptavidin/Biotin}
\subsection{Tutorial data}
We will use the 1STP crystal structure to start our calculations.
For the sake of simplicity, you can download prepared and solvated Amber topologies and structures based on 1STP by running
\begin{code}
git clone git@github.com:liedllab/gist-tutorial.git
\end{code}
This will download the tutorial files (including both the manuscript and the examples in the \software{code} folder).
\subsection{System Preparation and Equilibration}
When calculating solvation free energy differences between a complex and the corresponding monomers (the dissociated state), there are two options for dealing with configurational diversity in the structures.
\todo{Maybe Valentin knows a citation for that?}
One can choose to sample the complex and dissociated states individually, or to sample only one of them and assume that there is sufficient overlap in the sampled conformations.
%One can choose to sample only the complex, assuming that all relevant conformations of the dissociated state are also found in the complex.
%The probability distribution of the dissociated state can be obtained by reweighting the probabilities in the complex using the solvation free energies.
%Alternatively, one can sample the complex and dissociated states individually.
%This approach does not require an overlap between the conformational probabilities of the complex and dissociated states.
%However, it is prone to statistical errors due to uncertainties in the ensembles, and requires the free energy contributions from the internal degrees of freedom of the monomers (in both the complex and the monomer states) to be known. \todo{This paragraph does not fit here well. Maybe put it somewhere else.}

Here, we choose not to include any sampling of the conformational states. It is therefore important that the conformations of the monomers in the complex and dissociated states match exactly. We provide a script called \software{cpptraj\_remove\_mol.sh} that removes a single molecule from a topology and structure using \software{cpptraj} without altering the rest of the system. Use this to create monomer topologies of biotin and streptavidin.

\subsection{Running GIST}
\subsection{Visualizing $\dgsolv$}
\subsection{Contribution of Hydration to Binding}
